---
title: "Splines, part 05"
format: pptx
editor: source
---

## Most models will accept splines

-   Easy
    -   Linear regression
    -   Random effects regression models
-   Harder but not too difficult
    -   Generalized linear model
    -   Cox regression
-   Not at all helpful for data science models
    -   Random forests
    -   Deep neural nets

::: notes

Splines work with a variety of statistical models. They work very nicely with linear regresion and random effects regression models.

It takes a bit more work with the generalized linear model because of the nonlinear relationship with the outcome variable. With a bit of care, you will be fine.

Cox regression models also require a bit of care.

Many of the recent data science models do not work well with splines. The random forest model and deep neural nets, to name two, have their own way of modeling non-linearity and the spline functions would just get in the way.

Let me show how to use a spline in a logistic regression model with data from survival of passengers on the Titanic.

:::

## About the Titanic

-   Laregest passenger ship at its time
-   Maiden voyage in 1912
-   Hit an iceberg and sank
-   Detailed information on all passengers
    -   Survived (no/yes)
    -   Sex (female/male)
    -   Passenger class (1st, 2nd, 3rd)
    -   Age (2 months to 71 years)

::: notes

The Titanic was a large cruise ship, the biggest of its kind in 1912. It was thought to be unsinkable, but when it set sail from England to American in its maiden voyage, it struck an iceberg and sank, killing many of the passengers and crew. You can get fairly good data on the characteristics of passengers who died and compare them to those that survived. The data indicate a strong effect due to age and gender, representing a philosophy of "women and children first" that held during the boarding of life boats. Let's look at the effect of age on survival using a logistic regression model.

:::

## First few rows of data

```{r titanic-spline-02}
f2 <- "../data/titanic.txt"
ti0 <- 
  data.frame(
    read_tsv(
      file=f2,
      col_types="ccncn"))
names(ti0) <- tolower(names(ti0))
head(ti0)
```

## A few descriptive statistics

```{r titanic-spline-03}
summary(ti0$Age)
count(ti0, pclass)
count(ti0, sex)
count(ti0, survived)
```

## Boxplots

```{r titanic-spline-04}
ggplot(data=ti0, aes(factor(survived), age)) + 
  geom_boxplot()
```

::: notes

The boxplots reveal little differences between the ages of survivors and deaths. If something is going on, it is subtle.

:::

## Fit a linear model first.

```{r titanic-spline-05}
ti1 <- filter(ti0, !is.na(age))
linear_fit <- glm(
  survived ~ age,
  family=binomial,
  data=ti1)
tidy(linear_fit)
```

::: notes

There may be a downward trend in the odds of survival over time, but it is not statistically significant.

:::

## Plot the predicted values.

```{r titanic-spline-06}
age_sequence <- data.frame(
  age = seq(min(ti1$age), max(ti1$age), length=100))
age_sequence$prediction <- 
  predict(linear_fit, newdata=age_sequence, type="response")
ggplot(data=ti1, aes(age, survived)) +
  geom_point(pch=1) +
  geom_line(
    data=age_sequence,
    aes(age, prediction))
```

::: notes

The plot is linear, but you should really look at it on the log odds scale because the logistic regression model is linear in the log odds. You need to change the predict function from type="response" to type="link" to get predictions on a log odds scale.

:::

## Predicted values on the log odds scale

```{r titanic-spline-07}
age_sequence$prediction <- 
  predict(linear_fit, newdata=age_sequence, type="link")
ggplot(data=age_sequence, aes(age, prediction)) +
  geom_line()
```

## Now fit a spline function.

```{r titanic-spline-08}
spline_fit <- glm(
  survived ~ rcs(age),
  family=binomial,
  data=ti1)
tidy(spline_fit)
```

::: notes

The coefficients from the restricted cubic spline are pretty much uninterpretable. You have to visualize the spline graphically. First do this on the log odds scale to see how far from linear the spline fit is.

:::

## Predictions on a log odds scale

```{r titanic-spline-09}
age_sequence$prediction <- 
  predict(spline_fit, newdata=age_sequence, type="link")
ggplot(data=age_sequence, aes(age, prediction)) +
  geom_line()
```

## Predictions on the probability scale

```{r titanic-spline-10}
age_sequence$prediction <- 
  predict(spline_fit, newdata=age_sequence, type="response")
ggplot(data=ti1, aes(age, survived)) +
  geom_point(pch=1) +
  geom_line(
    data=age_sequence,
    aes(age, prediction))
```

::: notes

It looks like "women and children" first might actually be "women, children, and old people first".

:::