---
title: "See bottom of file for title"
format: pptx
params:
  topic_01: "Variety of regressions"
  topic_02: "Building cubic splines from scratch"
  topic_03: "B-splines, natural splines"
  topic_04: "How many knots and where to put them"
  topic_05: "Logistic regression example"
  topic_06: "Some code hints for R, SAS, Stata"
editor: source
---

```{r}
# Preliminaries, if needed

# Keep this chunk, even if no code, so that
# params is populated from the yaml header.
```

```{r}
#| label: setup
#| message: false
#| warning: false

# library(pspline)
library(readr)
library(rms)
library(tidyverse)
```


## Topics to be covered

-   What you will learn
    -   `r params$topic_01`
    -   `r params$topic_02`
    -   `r params$topic_03`
    -   `r params$topic_04`
    -   `r params$topic_05`
    -   `r params$topic_06`

{{< include _splines-01.qmd >}}

## Break #1

-   What you have learned
    -   `r params$topic_01`
-   What's coming next
    -   `r params$topic_02`

{{< include _splines-02.qmd >}}

## Break #2

-   What you have learned
    -   `r params$topic_02`
-   What's coming next
    -   `r params$topic_03`
    
{{< include _splines-03.qmd >}}

## Break #3

-   What you have learned
    -   `r params$topic_03`
-   What's coming next
    -   `r params$topic_04`
    
{{< include _splines-04.qmd >}}

## Break #4

-   What you have learned
    -   `r params$topic_04`
-   What's coming next
    -   `r params$topic_05`

{{< include _splines-05.qmd >}}

## Break #5

-   What you have learned
    -   `r params$topic_05`
-   What's coming next
    -   `r params$topic_06`
    
{{< include _splines-06.qmd >}}

## Summary

-   What you have learned
    -   `r params$topic_01`
    -   `r params$topic_02`
    -   `r params$topic_03`
    -   `r params$topic_04`
    -   `r params$topic_05`
    -   `r params$topic_06`

<!---
      You have to put the title down 
      here to override the titles of
      all the include files.
--->

---
title: "What are splines and how are they used?"
---